Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-10-02 morphology_paper.Rmd
general updates and additions. added customised reference list. added in new tables which can be found under the taables subfolder

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
fontsize: 10pt
output:
  word_document:
    reference_docx: frontiers_template.docx
  pdf_document: default
geometry: margin=1in
papersize: A4
tables: yes
bibliography: refs_morphology_paper.bib
---

```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

@@

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], nevertheless form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp forests [@Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Willis2003-xo; @anderson1997; @Dayton1985]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution [@Gorman2013-nj], abundance [@Cavanaugh2011-ib; @Dayton1998-xw], diversity [@Wing2007-tv; @Wernberg2008], composition [@Leliaert2000a; @Dayton1985; @Norderhaug2012-ks; @Harley2012], growth [@Cousens1982-ck] and productivity [@Pedersen2012-vr; @Dayton1998-xw; @Krumhansl2012] of kelps.

Given kelps are cool/temperate organisms which are vulernable to dislodgment under high wave exposure scenarios, changes in ocean temperature and an increase in storm frequency and magnitude pose the biggest direct threats to their survival. Warmer temperatures reduces the resilience of kelp individuals to perbutations [@wernberg2010], causes fragmentation through weakening tissues [@simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist which across a broad range of environments which is largely due to their morphological plasticity [@fowler2006; @reed2003]. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp].Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; Mia2013; @Dubi1995].

Although other factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main drivers of morphological adaptation across various kelp species [@bekkby2014; @wernberg2005; @Denny2002-ob; @Hurd2000; @Fowler-Walker2005; @Thomsen2004-yd; @mabin2013]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer sites and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Temperature affects important kelp physiological processes such as photosynthesis and respiration, which ultimately influences growth and productivity resulting in reduced size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002-ob; @Hurd2000; @Fowler-Walker2005; @Thomsen2004-yd]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata*  (C. Agardh) J. Agardh. across a broad geographic range, and showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

@@

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for quantifying wave exposure. Although @mabin2013 investigated kelp morphological characteristics using a 2 dimensional WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Furthermore no current studies have identified specific wave metrics in driving kelp morphological charactersitcs using advanced numerical model, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, shallow and deep water populations have not been considered using advanced numerical and statisical tools. The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology in two species of kelps around South Africa, as well as between deep and shallow populations of the same species. This will be achieved by initially understanding the variation in temperature and waves using an advanced numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological chracteristics of *E. maxima* and *L. pallida* using advanced statistical tools.

@@

Sites were chosen to represent an array of environmental gradients, as indicated in figure \ref{fig:map study sites}. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces and cover a cline in wave exposure and temperature. Study sites span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region [@Smit2013a].

@@

library(png)
library(grid)

@@

Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov]. Patterns are also evident in the wave data. Seasonal variations in significant wave height (Hs), peak period and wave direction are also present. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006], which False Bay is sheltered from [@Shipley1964; @Atkins1970-bx; @Dufois2012-vy]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012-ks; @Leliaert2000a]. It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.

@@

img <- readPNG("figures/kelp_diagram.png")
grid.raster(img)
```

@@

```{r Temperature data prep, message=FALSE, warning=FALSE, include=FALSE}
## Data wrangling and preparation
## Preparing data
# Complete lon/lats

@@

```{r Temperature variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for temperature variables 
positions_temp <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock", "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

@@

temp_boxplot <- ggplot(data = temp_ann_meta, aes(x = region, y = mean_temp, fill  = region)) + 
  geom_boxplot() + 
  stat_compare_means(label.y = 14.5 ,label.x = 1.5, label = "p.format") +
  xlab("") + 
  ylab("") +
  coord_flip() + 
  theme_bw()

```{r Temperature variables plot, echo=FALSE, warning=FALSE,fig.height=5,fig.width=8, fig.cap='\\label{fig:Temperature variables plot}Temperature variables at the collection sites around the Cape Peninsula are represented in plot A. Temperature variables include minimum represented by blue dots, mean represented by balc dots, maximum by red dots and whiskers standard deviation. Summary of temperature data for sites grouped by region are represented in plot B. See appendix for significance level guide.'}
ggarrange(temp_plot, temp_boxplot, common.legend = FALSE, ncol = 2, nrow = 1, labels = c("AUTO"), align = c("h"), legend = FALSE)
```

@@

```{r wave data prep, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
## Data wrangling and preparation
# Wave data 
# Load the loading functions
source("func/load.wave.R")
# Then load the data
fb_model <- load.model("data/wave_data/FB")
tb_model <- load.model("data/wave_data/TB")
he_model <- load.model("data/wave_data/HE")
sh_model <- load.model("data/wave_data/SH")

@@

```{r wave variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for wave variables
#Annual wave plots
#combine with meta data for easier plotting
wave_ann_meta <- merge(meta_all, wave_ann, by = "site")
wave_ann_meta$site <- as.character(wave_ann_meta$site)#make character to ensure no issues arise for site and region names

@@

wave_boxplot_1 <- ggplot(data = wave_ann_meta, aes(x = region, y = hs_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y =1.4,label.x = 1.5, label = "p.format") + 
  xlab("") + 
  ylab("") + 
  coord_flip() + 
  theme_bw()

wave_boxplot_2 <- ggplot(data = wave_ann_meta, aes(x = region, y = tp_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y = 10, label.x = 1.5, label = "p.format") + 
  xlab("") + 
  ylab("") +
  ylim(8,12) + 
  coord_flip() + 
  theme_bw()

@@

```{r wave variable plot, warning = FALSE, echo=FALSE, message = FALSE, include= TRUE,fig.height=7,fig.width=8, fig.cap='\\label{fig:wave variable plot}In plot A1 and A2 wave variables at each collection site around the Western Cape coast are presented. Mean variables are represented by dots and standard deviation by whiskers. Blue dots represent sites grouped to the western category and red dots represent sites grouped to the eastern category.'}
ggarrange(wave_plot_A,wave_boxplot_1, wave_plot_B, wave_boxplot_2, common.legend = TRUE, labels = c("A", "C", "B", "D"))
```

@@

img_wave_rose <- readPNG("figures/wave_rose.png")
grid.raster(img_wave_rose)
```

@@

```{r wave power map, echo=FALSE, eval= FALSE, fig.cap='\\label{fig:wave power map}Wave power map'}

library(png)
library(grid)

img_wavepower <- readPNG("figures/CapePen_Total.PNG")
legend_wave_power <- readPNG("figures/CapePen_Legend.PNG")
grid.raster(img_wavepower,legend_wave_power)
```

@@

```{r FB spectrum, echo=FALSE, eval= FALSE, fig.cap='\\label{fig:FB spectrum}Wave power map'}

library(png)
library(grid)

img_fb <- readPNG("figures/falsebay_spectrum.png")
grid.raster(img_fb)
```

@@

```{r laminaria processing, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
#custom temperature categories position on x-axis
pos_temp <- c("Cool temperate", "Warm temperate")

#temperature and lamanaria  
LT1 <- ggplot(data = lam_data, aes(x = temp_regime, y = laminaWeight, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp) + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Lamina weight") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT2 <- ggplot(data = lam_data, aes(x = temp_regime, y = laminaLength, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Lamina length") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT3 <- ggplot(data = lam_data, aes(x = temp_regime, y = laminaThickness, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp) + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Lamina thickness") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT4 <- ggplot(data = lam_data, aes(x = temp_regime, y = stipeMass, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Stipe mass (g)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT5 <- ggplot(data = lam_data, aes(x = temp_regime, y = stipeLength, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Stipe length (cm)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT6 <- ggplot(data = lam_data, aes(x = temp_regime, y = stipeDiameter, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Stipe diameter (cm)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT7 <- ggplot(data = lam_data, aes(x = temp_regime, y = digits, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("No. of digits") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT8 <- ggplot(data = lam_data, aes(x = temp_regime, y = thallusMass, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp) + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Thallus mass (g)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

>>>>

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
fontsize: 10pt
output:
  word_document:
    reference_docx: frontiers_template.docx
  pdf_document: 
    fig_caption: yes
geometry: margin=1in
papersize: A4
tables: yes
bibliography: refs_morphology_paper.bib
---

```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(fig.align = "center")
```

@@

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp forests [@Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Willis2003-xo; @anderson1997; @Dayton1985]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution [@Gorman2013-nj], abundance [@Cavanaugh2011-ib; @Dayton1998-xw], diversity [@Wing2007-tv; @Wernberg2008], composition [@Leliaert2000a; @Dayton1985; @Norderhaug2012-ks; @Harley2012], growth [@Cousens1982-ck] and productivity [@Pedersen2012-vr; @Dayton1998-xw; @Krumhansl2012] of kelps.

Given kelps are cool/temperate organisms which are vulernable to dislodgment under high wave exposure scenarios, changes in ocean temperature and an increase in storm frequency and magnitude pose the biggest direct threats to their survival. Warmer temperatures reduces the resilience of kelp individuals to perbutations [@wernberg2010], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @wernberg2010]. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity, and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; Mia2013; @Dubi1995].

Although other factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main drivers of morphological adaptation across various kelp species [@bekkby2014; @Wernberg2005c; @Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004; @mabin2013]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Temperature affects important kelp physiological processes such as photosynthesis and respiration, which ultimately influences growth and productivity resulting in reduced size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata*  (C. Agardh) J. Agardh. across a broad geographic range, and showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

@@

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for characterising the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Furthermore no current studies have identified specific wave metrics in driving kelp morphological charactersitcs using advanced numerical model, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, shallow and deep water populations have not been considered using advanced numerical and statisical tools. The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology in two species of kelps around South Africa, as well as between deep and shallow populations of the same species. The current study will also aim to identify specific temperature and wave metrics which drive kelp morphological characteristics.  This will be achieved by initially understanding the variation in temperature and waves using an advanced numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological chracteristics of *E. maxima* and *L. pallida* using advanced statistical tools.

@@

Sites were chosen to represent an array of environmental gradients, as indicated in figure \ref{fig:map study sites}. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, lie within separate marine provinces, and span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region [@Smit2013a].

@@

library(png)
library(grid)
library(gridExtra)

@@

The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient rich water to coastal areas and therefore induces a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape peninsula, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay which is shielded from the dominating wind and swell the Cape Peninsula is subject to, with the exception being the winter months where upwelling becomes more favourable for the region due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov].

@@

schematic <- readPNG("figures/kelp_diagram.png")
grid.raster(schematic)
```

@@

```{r Temperature data prep, message=FALSE, warning=FALSE, include=FALSE, eval=TRUE}
## Data wrangling and preparation
## Preparing data
# Complete lon/lats

@@

```{r Temperature variable processing, message=FALSE, warning=FALSE, include=FALSE, eval=TRUE}
### Plots for temperature variables 
positions_temp <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock", "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

@@

temp_boxplot <- ggplot(data = temp_ann_meta, aes(x = region, y = mean_temp, fill  = region)) + 
  geom_boxplot() + 
  stat_compare_means(label.x = 2.5, label.y = 12.5, label = "p.format") +
  xlab("") + 
  ylab("") +
  coord_flip() + 
  theme_bw()
```

```{r Temperature variables plot, echo=FALSE, warning=FALSE, eval=TRUE,fig.height=5,fig.width=8, fig.cap='\\label{fig:Temperature variables plot}Temperature variables at the collection sites around the Cape Peninsula are represented in plot A. Temperature variables include minimum represented by blue dots, mean represented by balc dots, maximum by red dots and whiskers standard deviation. Summary of temperature data for sites grouped by region are represented in plot B. See appendix for significance level guide.'}
ggarrange(temp_plot, temp_boxplot, common.legend = FALSE, ncol = 2, nrow = 1, labels = c("AUTO"), align = c("h"), legend = FALSE)
```

@@

```{r wave data prep, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE, eval=TRUE}
## Data wrangling and preparation
# Wave data 
# Load the loading functions
source("func/load.wave.R")
# Then load the data
fb_model <- load.model("data/wave_data/FB")
tb_model <- load.model("data/wave_data/TB")
he_model <- load.model("data/wave_data/HE")
sh_model <- load.model("data/wave_data/SH")

@@

```{r wave variable processing, message=FALSE, warning=FALSE, include=FALSE, eval=TRUE}
### Plots for wave variables
#Annual wave plots
#combine with meta data for easier plotting
wave_ann_meta <- merge(meta_all, wave_ann, by = "site")
wave_ann_meta$site <- as.character(wave_ann_meta$site)#make character to ensure no issues arise for site and region names

@@

wave_boxplot_1 <- ggplot(data = wave_ann_meta, aes(x = region, y = hs_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y =0.5 ,label.x = 2.5, label = "p.format") + 
  xlab("") + 
  ylab("") + 
  coord_flip() + 
  theme_bw()

wave_boxplot_2 <- ggplot(data = wave_ann_meta, aes(x = region, y = tp_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y = 8.1, label.x = 2.5, label = "p.format") + 
  xlab("") + 
  ylab("") +
  ylim(8,12) + 
  coord_flip() + 
  theme_bw()

@@

```{r wave variable plot, warning = FALSE, echo=FALSE, message = FALSE, include= TRUE, eval=TRUE,fig.height=7,fig.width=8, fig.cap='\\label{fig:wave variable plot}In plot A1 and A2 wave variables at each collection site around the Western Cape coast are presented. Mean variables are represented by dots and standard deviation by whiskers. Blue dots represent sites grouped to the western category and red dots represent sites grouped to the eastern category.'}
ggarrange(wave_plot_A,wave_boxplot_1, wave_plot_B, wave_boxplot_2, common.legend = TRUE, labels = c("A", "C", "B", "D"))
```

@@

library(png)
library(grid)
library(gridExtra)

img_wave_rose <- readPNG("figures/wave_rose.png")
grid.raster(img_wave_rose)

@@

```{r wave power map, echo=FALSE, eval= TRUE, fig.cap='\\label{fig:wave power map}Wave power map'}

library(png)
library(grid)
library(gridExtra)

wp_1 <-  rasterGrob(as.raster(readPNG("figures/CapePen_Total.PNG")), interpolate = FALSE)
wp_2 <-  rasterGrob(as.raster(readPNG("figures/CapePen_Legend.PNG")), interpolate = FALSE)
grid.arrange(wp_1, wp_2, ncol = 2, widths = c(2,1))

@@

```{r FB spectrum, echo=FALSE, eval= TRUE, fig.cap='\\label{fig:FB spectrum}False Bay wave spectrum'}

library(png)
library(grid)
library(jpeg)

img_fb <- readJPEG("figures/falsebay_spectrum.jpg")
grid.raster(img_fb)

@@

```{r laminaria processing, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE, eval=TRUE}
#custom temperature categories position on x-axis
pos_temp <- c("Cool temperate", "Warm temperate")

#temperature and lamanaria  
LT1 <- ggplot(data = lam_data, aes(x = temp_regime, y = laminaWeight, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp) + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Lamina weight") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT2 <- ggplot(data = lam_data, aes(x = temp_regime, y = laminaLength, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Lamina length") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT3 <- ggplot(data = lam_data, aes(x = temp_regime, y = laminaThickness, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp) + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Lamina thickness") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT4 <- ggplot(data = lam_data, aes(x = temp_regime, y = stipeMass, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Stipe mass (g)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT5 <- ggplot(data = lam_data, aes(x = temp_regime, y = stipeLength, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Stipe length (cm)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT6 <- ggplot(data = lam_data, aes(x = temp_regime, y = stipeDiameter, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("Stipe diameter (cm)") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")

LT7 <- ggplot(data = lam_data, aes(x = temp_regime, y = digits, fill = temp_regime)) +
  scale_x_discrete(limits = pos_temp)  + 
  geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = FALSE, show.legend = FALSE) + 
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x  = 0.6, label.y.npc = c('top')) + 
  xlab("") + 
  ylab("No. of digits") +
  scale_fill_discrete(name="Temperature regime") +
  theme_bw() +
  theme(axis.text.x = element_blank()) + 
  xlab("")
